---
title: "Presentation - Don't commit Suicide"
author: "Marckenrold Cadet & Alexander Burkhart"
date: "`r Sys.Date()`"
output: 
   html_notebook:
       toc: true 
       toc_float: true
       toc_depth: 2
       theme: united
       highlight: tango
---

```{r}
library(dplyr)
library(readr)
library(tidyverse)
library(ggplot2)
library(highcharter)
library(magrittr)
```
# EDA
## Data loading
```{r}
data <- read_csv("https://raw.githubusercontent.com/Alexburk93/Data_Wrangling_EDA/main/data/suicide_analysis_2.csv")
                data %>% sample_n(19)
```
## Renaming variables and drop unwanted columns
```{r}
# renaming the columns 
data <- data %>%
  rename(`Country_name` = `Country name`,
         `Suicide_Rate` = `Age-standardized suicide rate - Sex: both sexes`,
         `Life_ladder` = `Life Ladder`,
         `Social_support` = `Social support`,
         `Life_expectancy` = `Healthy life expectancy at birth`,
         `Freedom_choices` = `Freedom to make life choices`,
         `Corruption` = `Perceptions of corruption`)
```

```{r}
data
```
```{r}
data
```


```{r}
# drop columns 
# remove the original 'Time' column
data <- select(data, -Variable, -VAR, -MEA, -`Unit Code`)  
data
```

## Data exploration
```{r}
head(data)
```

```{r}
# change names
names(data) <- make.names(names(data))
```

```{r}
# dimensions of the dataframe
nrow(data)
ncol(data)
dim(data)
```

```{r}
# check the structure of the object
str(data)
```
```{r}
# look at columns 6, 7 and 10
head(data[ , c(2, 4:6, 12, 15)])
```

```{r}
# look at columns 6, 7 and 10
tail(data[ , c(1, 3, 9)])
```

```{r}
table(data$Year)
```


```{r}
data %>% 
  select(Country_name) %>% 
  unique() %>% 
  nrow()
```

```{r}
unique(data$Country_name)
```

```{r}
unique(data$Year)
```

## Interactive maps
```{r}
# Set highcharter options for tooltip decimals
options(highcharter.tooltip.valueDecimals = 2)

# Create highcharter map visualization
hc <- highchart() %>%
  hc_add_series_map(
    worldgeojson, data, value = "GDP", 
    joinBy = c('name', 'Country_name'),
    name = "GDP (current US$)"
  )  %>% 
  hc_colorAxis(stops = color_stops()) %>% 
  hc_title(text = "World Map") %>% 
  hc_subtitle(text = "GDP in current US$")

hc
```

```{r}
# Set highcharter options for tooltip decimals
options(highcharter.tooltip.valueDecimals = 2)

# Create map visualizations for each variable
hc_life_expectancy <- highchart() %>%
  hc_add_series_map(
    worldgeojson, data, 
    value = "Life_expectancy", 
    joinBy = c('name', 'Country_name'),
    name = "Life Expectancy"
  ) %>%
  hc_colorAxis(stops = color_stops()) %>%
  hc_title(text = "World Map") %>%
  hc_subtitle(text = "Life Expectancy")

hc_suicide_rates <- highchart() %>%
  hc_add_series_map(
    worldgeojson, data, 
    value = "Suicide_Rate", 
    joinBy = c('name', 'Country_name'),
    name = "Suicide Rates"
  ) %>%
  hc_colorAxis(stops = color_stops()) %>%
  hc_title(text = "World Map") %>%
  hc_subtitle(text = "Suicide Rate")

hc_corruption <- highchart() %>%
  hc_add_series_map(
    worldgeojson, data, 
    value = "Corruption", 
    joinBy = c('name', 'Country_name'),
    name = "Corruption"
  ) %>%
  hc_colorAxis(stops = color_stops()) %>%
  hc_title(text = "World Map") %>%
  hc_subtitle(text = "Corruption")

# Display the map visualizations
list(hc_life_expectancy, hc_suicide_rates, hc_corruption)

```

# AVG GDP over years
## Calculation 
```{r}
avg_gdp_per_year <- data %>% 
  group_by (`Year`) %>% 
  summarise(avg_gpd = mean(`GDP`))

avg_gdp_per_year
```
## Plot 
```{r}
ggplot(avg_gdp_per_year, aes(x = Year, y = avg_gpd)) +
  geom_line(color = "blue") +  
  labs(title = "Average GDP Over Time worldwide",
       x = "Year",
       y = "GDP") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_gdp_per_year$Year), max(avg_gdp_per_year$Year), by = 1))
```
# AVG Happiness over years
## Calculation 
```{r}
avg_happiness_per_year <- data %>% 
  group_by (`Year`) %>% 
  summarise(avg_happinnes = mean(`Life_ladder`, na.rm = T))

avg_happiness_per_year
```
## Plot 

```{r}
ggplot(avg_happiness_per_year, aes(x = Year, y = avg_happinnes)) +
  geom_line(color = "blue") +  
  labs(title = "Average Happiness Over Time worldwide",
       x = "Year",
       y = "Happiness") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_happiness_per_year$Year), max(avg_happiness_per_year$Year), by = 1))
```

# AVG Suicide Rates over years
## Calculation 
```{r}
avg_Suicide_Rate_per_year <- data %>% 
  group_by (`Year`) %>% 
  summarise(avg_Suicide_Rate = mean(`Suicide_Rate`, na.rm = T))

avg_Suicide_Rate_per_year
```
## Plot 

```{r}
ggplot(avg_Suicide_Rate_per_year, aes(x = Year, y = avg_Suicide_Rate)) +
  geom_line(color = "blue") +  
  labs(title = "Average Suicide Rate Over Time worldwide",
       x = "Year",
       y = "Suicide Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_Suicide_Rate_per_year$Year), max(avg_Suicide_Rate_per_year$Year), by = 1))
```
# AVG Bankruptcies over years
## Calculation 
```{r}
avg_Bankruptcies_per_year <- data %>% 
  group_by (`Year`) %>% 
  summarise(avg_Bankruptcies = mean(`Value`, na.rm = T))

avg_Bankruptcies_per_year
```
## Plot 

```{r}
ggplot(avg_Bankruptcies_per_year, aes(x = Year, y = avg_Bankruptcies)) +
  geom_line(color = "blue") +  
  labs(title = "Average Bankruptcies Over Time worldwide",
       x = "Year",
       y = "Bankruptcies") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_Bankruptcies_per_year$Year), max(avg_Bankruptcies_per_year$Year), by = 1))
```

# Plot Average GDP and Average Suicide Rate over years
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_gdp_per_year$avg_gpd) / max(avg_Suicide_Rate_per_year$avg_Suicide_Rate)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_gdp_per_year, aes(x = Year, y = avg_gpd), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_Suicide_Rate_per_year, aes(x = Year, y = avg_Suicide_Rate * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average GDP and Suicide Rate Over Time",
       x = "Year",
       y = "Average GDP") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_gdp_per_year$Year, avg_Suicide_Rate_per_year$Year)), 
                                  max(c(avg_gdp_per_year$Year, avg_Suicide_Rate_per_year$Year)), by = 1))

```

# Plot Average Happinness and Average Suicide Rate over years
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_happiness_per_year$avg_happinnes) / max(avg_Suicide_Rate_per_year$avg_Suicide_Rate)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_happiness_per_year, aes(x = Year, y = avg_happinnes), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_Suicide_Rate_per_year, aes(x = Year, y = avg_Suicide_Rate * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average Happinness and Suicide Rate Over Time",
       x = "Year",
       y = "Average Happinness") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_happiness_per_year$Year, avg_Suicide_Rate_per_year$Year)), 
                                  max(c(avg_happiness_per_year$Year, avg_Suicide_Rate_per_year$Year)), by = 1))

```

# Plot Average Bankruptcies and Average Suicide Rate over years
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_Bankruptcies_per_year$avg_Bankruptcies) / max(avg_Suicide_Rate_per_year$avg_Suicide_Rate)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_Bankruptcies_per_year, aes(x = Year, y = avg_Bankruptcies), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_Suicide_Rate_per_year, aes(x = Year, y = avg_Suicide_Rate * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average Bankruptcies and Suicide Rate Over Time",
       x = "Year",
       y = "Average Bankruptcies") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_Bankruptcies_per_year$Year, avg_Suicide_Rate_per_year$Year)), 
                                  max(c(avg_Bankruptcies_per_year$Year, avg_Suicide_Rate_per_year$Year)), by = 1))

```

# Comparisions

## Two Happiest and Two unhappiest countries vs average suicide rate
```{r}
avg_Suicide_Rate_per_country = data %>%
  group_by(Country_name) %>%
  summarise(avg_suicide_rate = mean(Suicide_Rate, na.rm = TRUE)) %>%
  arrange(avg_suicide_rate) %>% 
  mutate(Row_Number = row_number())

avg_Suicide_Rate_per_country

avg_happiness_per_country <- data %>%
  group_by(Country_name) %>%
  summarise(avg_happiness = mean(Life_ladder, na.rm = TRUE)) %>%
  arrange(desc(avg_happiness))

least_happy =  tail(avg_happiness_per_country, 2)
most_happy = head(avg_happiness_per_country, 2)


avg_Suicide_Rate_per_country %>% 
  filter(Country_name %in% least_happy$Country_name)

# Interpretation: Japan and South Africa are two very unhappy countries. And they also have a high suicide rate

avg_Suicide_Rate_per_country %>% 
  filter(Country_name %in% most_happy$Country_name)

# Interpretation: Finland is a the second most happy country. But is still on place 16/19 when it comes to suicides

```

## Two most wealthy countries and two most poor countries vs average suicide rate
```{r}
avg_gdp_per_country <- data %>% 
  group_by (`Country_name`) %>% 
  summarise(avg_gpd = mean(`GDP`)) %>% 
  arrange(desc(avg_gpd))

least_gdp =  tail(avg_gdp_per_country, 2)
most_gdp = head(avg_gdp_per_country, 2)

avg_Suicide_Rate_per_country %>% 
  filter(Country_name %in% least_gdp$Country_name)

# Interpretation: New Zealand and Iceland are two very small countries.  

avg_Suicide_Rate_per_country %>% 
  filter(Country_name %in% most_gdp$Country_name)

# Interpretation: The two biggest economies are based on the lower end of suicide rates
```


## Two most bankcuptcies and two least bankcuptcies countries vs average suicide rate
```{r}

avg_Bankruptcies_per_year <- data %>% 
  group_by (`Country_name`) %>% 
  summarise(avg_bankruptcies = mean(`Value`, na.rm = T)) %>% 
  arrange(desc(avg_bankruptcies))

least_bank =  tail(avg_Bankruptcies_per_year, 2)
most_bank = head(avg_Bankruptcies_per_year, 2)

avg_Suicide_Rate_per_country %>% 
  filter(Country_name %in% least_bank$Country_name)

# Interpretation: Bankruptcies don't have an influence on suicide rates

avg_Suicide_Rate_per_country %>% 
  filter(Country_name %in% most_bank$Country_name)

# Interpretation: Bankruptcies don't have an influence on suicide rates
```
# In depth analysis Germany
## Data preperation for Germany
```{r}
# Prepare data for only Germany
germany_data = data %>% 
  filter(Country_name == "Germany")
```

## Plot GDP Germany
```{r}
# Plot Germany GDP over Years
avg_gdp_year_germany = germany_data %>% 
  group_by(Year) %>% 
  summarise(avg_gdp = mean(GDP)) 


ggplot(avg_gdp_year_germany, aes(x = Year, y = avg_gdp)) +
  geom_line(color = "blue") +  
  labs(title = "Average GPD Over Time - Germany",
       x = "Year",
       y = "GDP") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_gdp_year_germany$Year), max(avg_gdp_year_germany$Year), by = 1))

```

## Plot Suicide Rate Germany
```{r}
avg_suicide_year_germany = germany_data %>% 
  group_by(Year) %>% 
  summarise(avg_suicide = mean(Suicide_Rate)) 


ggplot(avg_suicide_year_germany, aes(x = Year, y = avg_suicide)) +
  geom_line(color = "blue") +  
  labs(title = "Average Suicide Rate Over Time - Germany",
       x = "Year",
       y = "Suicide Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_suicide_year_germany$Year), max(avg_suicide_year_germany$Year), by = 1))
```
## Plot Bankruptcies Rate Germany
```{r}
avg_bank_year_germany = germany_data %>% 
  group_by(Year) %>% 
  summarise(avg_bank = mean(Value)) 


ggplot(avg_bank_year_germany, aes(x = Year, y = avg_bank)) +
  geom_line(color = "blue") +  
  labs(title = "Average bankruptcies Over Time - Germany",
       x = "Year",
       y = "Bankruptcies") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_bank_year_germany$Year), max(avg_bank_year_germany$Year), by = 1))
```
## Plot Happiness Rate Germany
```{r}
avg_happiness_year_germany = germany_data %>% 
  group_by(Year) %>% 
  summarise(avg_happy = mean(Life_ladder)) 


ggplot(avg_happiness_year_germany, aes(x = Year, y = avg_happy)) +
  geom_line(color = "blue") +  
  labs(title = "Average Happiness Over Time - Germany",
       x = "Year",
       y = "Happiness") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_happiness_year_germany$Year), max(avg_happiness_year_germany$Year), by = 1))
```
# Plot Average GDP and Average Suicide Rate over years - Germany
```{r}
avg_suicide_year_germany$avg_suicide
# Finding the ratio for scaling the second axis
ratio <- max(avg_gdp_year_germany$avg_gdp) / max(avg_suicide_year_germany$avg_suicide)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_gdp_year_germany, aes(x = Year, y = avg_gdp), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_suicide_year_germany, aes(x = Year, y = avg_suicide * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average GDP and Suicide Rate Over Time",
       x = "Year",
       y = "Average GDP") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_gdp_year_germany$Year, avg_suicide_year_germany$Year)), 
                                  max(c(avg_gdp_year_germany$Year, avg_suicide_year_germany$Year)), by = 1))
```

# Plot Average Happinness and Average Suicide Rate over years - Germany
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_happiness_year_germany$avg_happy) / max(avg_suicide_year_germany$avg_suicide)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_happiness_year_germany, aes(x = Year, y = avg_happy), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_suicide_year_germany, aes(x = Year, y = avg_suicide * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average Happinness and Suicide Rate Over Time",
       x = "Year",
       y = "Average Happinness") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_happiness_year_germany$Year, avg_suicide_year_germany$Year)), 
                                  max(c(avg_happiness_year_germany$Year, avg_suicide_year_germany$Year)), by = 1))

```
# Plot Average Bankruptcies and Average Suicide Rate over years
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_bank_year_germany$avg_bank) / max(avg_suicide_year_germany$avg_suicide)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_bank_year_germany, aes(x = Year, y = avg_bank), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_suicide_year_germany, aes(x = Year, y = avg_suicide * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average Bankruptcies and Suicide Rate Over Time",
       x = "Year",
       y = "Average Bankruptcies") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_bank_year_germany$Year, avg_suicide_year_germany$Year)), 
                                  max(c(avg_bank_year_germany$Year, avg_suicide_year_germany$Year)), by = 1))

```


# In depth analysis South Africa
## Data preperation for South Africa
```{r}
# Prepare data for only Germany
SA_data = data %>% 
  filter(Country_name == "South Africa")
```

## Plot GDP SA
```{r}
# Plot SA GDP over Years
avg_gdp_year_SA = SA_data %>% 
  group_by(Year) %>% 
  summarise(avg_gdp = mean(GDP)) 



ggplot(avg_gdp_year_SA, aes(x = Year, y = avg_gdp)) +
  geom_line(color = "blue") +  
  labs(title = "Average GPD Over Time - SA",
       x = "Year",
       y = "GDP") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_gdp_year_SA$Year), max(avg_gdp_year_SA$Year), by = 1))

```
## Plot Suicide Rate SA
```{r}
avg_suicide_year_SA = SA_data %>% 
  group_by(Year) %>% 
  summarise(avg_suicide = mean(Suicide_Rate)) 


ggplot(avg_suicide_year_SA, aes(x = Year, y = avg_suicide)) +
  geom_line(color = "blue") +  
  labs(title = "Average Suicide Rate Over Time - SA",
       x = "Year",
       y = "Suicide Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_suicide_year_SA$Year), max(avg_suicide_year_SA$Year), by = 1))
```

## Plot Happiness Rate SA
```{r}
avg_happiness_year_SA = SA_data %>% 
  group_by(Year) %>% 
  summarise(avg_happy = mean(Life_ladder)) 


ggplot(avg_happiness_year_SA, aes(x = Year, y = avg_happy)) +
  geom_line(color = "blue") +  
  labs(title = "Average Happiness Over Time - SA",
       x = "Year",
       y = "Happiness") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_happiness_year_SA$Year), max(avg_happiness_year_SA$Year), by = 1))
```
## Plot Bankruptcies Rate SA
```{r}
avg_bank_year_sa = SA_data %>% 
  group_by(Year) %>% 
  summarise(avg_bank = mean(Value)) 


ggplot(avg_bank_year_sa, aes(x = Year, y = avg_bank)) +
  geom_line(color = "blue") +  
  labs(title = "Average bankruptcies Over Time - SA",
       x = "Year",
       y = "Bankruptcies") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(avg_bank_year_sa$Year), max(avg_bank_year_sa$Year), by = 1))
```

# Plot Average GDP and Average Suicide Rate over years - SA
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_gdp_year_SA$avg_gdp) / max(avg_suicide_year_SA$avg_suicide)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_gdp_year_SA, aes(x = Year, y = avg_gdp), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_suicide_year_SA, aes(x = Year, y = avg_suicide * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average GDP and Suicide Rate Over Time - SA",
       x = "Year",
       y = "Average GDP") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_gdp_year_SA$Year, avg_suicide_year_SA$Year)), 
                                  max(c(avg_gdp_year_SA$Year, avg_suicide_year_SA$Year)), by = 1))
```
# Plot Average Bankruptcies and Average Suicide Rate over years
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_bank_year_sa$avg_bank) / max(avg_suicide_year_SA$avg_suicide)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_bank_year_sa, aes(x = Year, y = avg_bank), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_suicide_year_SA, aes(x = Year, y = avg_suicide * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average Bankruptcies and Suicide Rate Over Time - SA",
       x = "Year",
       y = "Average Bankruptcies") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_bank_year_sa$Year, avg_suicide_year_SA$Year)), 
                                  max(c(avg_bank_year_sa$Year, avg_suicide_year_SA$Year)), by = 1))
```
# Plot Average Happinness and Average Suicide Rate over years - SA
```{r}
# Finding the ratio for scaling the second axis
ratio <- max(avg_happiness_year_SA$avg_happy) / max(avg_suicide_year_SA$avg_suicide)

# Creating the base plot
ggplot() +
  # Adding the bar plot for GDP
  geom_bar(data = avg_happiness_year_SA, aes(x = Year, y = avg_happy), stat = "identity", fill = "skyblue", width = 0.2) +
  # Adding the line plot for Average Happiness
  geom_line(data = avg_suicide_year_SA, aes(x = Year, y = avg_suicide * ratio), color = "red", size = 1.5) +
  # Enhancing the plot
  labs(title = "Average Happinness and Suicide Rate Over Time - SA",
       x = "Year",
       y = "Average Happinness") +
  scale_y_continuous(sec.axis = sec_axis(~ . / ratio, name = "Average Suicide Rate")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_continuous(breaks = seq(min(c(avg_happiness_year_SA$Year, avg_suicide_year_SA$Year)), 
                                  max(c(avg_happiness_year_SA$Year, avg_suicide_year_SA$Year)), by = 1))
```