---
title: "HW 5 - DW and EDA (Project Progress Part - 01)"
author: "Alexander Burkhart & Marckenrold Cadet"
date: "2024-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Suicide is the deliberate act of ending one's own life, often stemming from various mental disorders such as depression, bipolar disorder, autism, schizophrenia, and personality disorders, as well as external stressors like financial struggles, academic pressures, relationship issues, or experiences of harassment and bullying. Factors like substance abuse, including alcoholism and benzodiazepine use, also contribute to this tragic outcome. Previous suicide attempts significantly elevate the risk of future attempts.

Efforts to prevent suicide involve a multifaceted approach, including restricting access to common methods like firearms, drugs, and poisons, addressing mental health issues and substance abuse, responsible media reporting on suicides, and fostering better economic conditions. Despite the widespread availability of crisis hotlines, their effectiveness remains inadequately researched.

The prevalence and methods of suicide vary across countries, often influenced by the accessibility of lethal means. Hanging, pesticide ingestion, poisoning, and firearms are among the most commonly used methods. Globally, suicides claim over 700,000 lives annually, ranking suicide as the 10th leading cause of death worldwide.

Approximately 1.5% of people die by suicide, translating to roughly 12 per 100,000 individuals each year. Men are more likely to die by suicide than women, with rates ranging from 1.5 times higher in developing countries to 3.5 times higher in developed ones. Financial strain often exacerbates the risk of suicide. [1]

Suicide is not confined to high-income countries; it is a global issue affecting all regions. Surprisingly, over 77% of suicides occur in low- and middle-income countries. However, high-income countries exhibit the highest age-standardized suicide rates. In ongoing research across 19 countries, we explore correlations between GDP, bankruptcy rates, happiness levels, and suicide rates. Understanding these intricate relationships is pivotal for policymakers, researchers, and stakeholders to devise effective interventions promoting mental well-being and economic resilience. Leveraging multidimensional datasets and advanced analytical methods, our project aims to illuminate these connections for the betterment of societies worldwide.

### Dataset Description
```{r}
# Load libraries
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(stringr)
library(readxl)
library(httr)
```

### Exploration of the Bankruptcies Dataset
```{r}

# Load dataset
bankruptcies <- read_csv("https://raw.githubusercontent.com/Alexburk93/Data_Wrangling_EDA/main/data/raw_data/Bankruptcies_2011-2020.csv")

# Display structure of dataset
str(bankruptcies)

# Show the first few rows 
head(bankruptcies)

# Summary statistics of numerical attributes
summary(bankruptcies)

# Summary statistics of categorical attribute
table(bankruptcies$Country)

# Check for missing values
sum(is.na(bankruptcies))

# Removing NA's columns and Time (Duplicate)  
data <- select(bankruptcies, -`Flag Codes`, -Flags, -`Reference Period Code`, -`Reference Period`, -TIME)
data %>% 
  sample_n(10)

# Display the structure of the modified dataset
str(data)

# Check the dimension of data
dim(data)

```

For this project, we are only focusing on information from 2011 to 2020. And to minimize the workload, we are analyzing 19 countries. 
```{r}
# Create new columns 'Quarter' and 'Year' from 'Time'
new_data <- mutate(data,
                   Quarter = str_sub(Time, 1, 2), # Extract Quarter
                   Year = as.numeric(str_sub(Time, 4))) # Extract Year and convert to numeric

# remove the original 'Time' column
new_data <- select(new_data, -Time)  
new_data

# Get date range
date_range <- range(new_data$Year)

# Count unique countries
n_countries <- length(unique(new_data$Country))

# Print the results
n_countries
date_range

# Only need data from 2011-2020, excluding 2021
bankruptcy_data <- filter(new_data, Year != 2021)
new_data

# Get date range of new updated dataset
date_range <- range(bankruptcy_data$Year)

# Extract start and end years from the range
start_year <- date_range[1]
end_year <- date_range[2]

# Count unique countries
n_countries <- length(unique(bankruptcy_data$Country))

# Print the results
cat("Number of country: ", n_countries, "\n") 
cat("Date range from: ", start_year, " to ", end_year, "\n")
```

### Exploring GDP dataset & analyze foundings
```{r}
# Load dataset 
my_data <- read_csv("https://raw.githubusercontent.com/Alexburk93/Data_Wrangling_EDA/main/data/raw_data/GDP_Data/GDP_capita_1960_2022.csv")
my_data
```

The Years are in the header, using `pivot_longer()` to convert into two column Year and GDP
```{r}
# Pivot the data from wide to long format
joy <- pivot_longer(my_data, 
                         cols = -c(`Country Name`),
                         names_to = "Year", 
                         values_to = "GDP")

# Show the first few rows

```


Perfom `summary stastics`, check for `NA's`, dimension `dim()`, etc. 
```{r}
# Display structure of dataset
str(joy)

# Show the first few rows 
head(joy)

# Summary statistics of numerical attributes
summary(joy)

# Summary statistics of categorical attribute
table(joy$`Indicator Code`)

# Check for missing values
sum(is.na(joy))

# Check the dimension of data
dim(joy)
```

Filter the data to only show the 19 countries and the date range the project is focused on.

```{r}
# Define the list of 19 countries
countries_of_interest <- c("Australia", "Belgium", "Brazil", "Canada", "Denmark",
                           "Finland", "France", "Germany", "Iceland", "Italy",
                           "Japan", "Netherlands", "New Zealand", "Norway", 
                           "South Africa", "Spain", "Sweden", "United Kingdom", 
                           "United States")

# Filter the data for the 19 countries
gdp <- joy %>%
  filter(`Country Name` %in% countries_of_interest)

# Define the date range of focus
start_year <- 2011
end_year <- 2020

# Filter the data for the date range
gdp <- gdp %>%
  filter(as.numeric(Year) >= start_year & as.numeric(Year) <= end_year)

# Show the filtered data
gdp %>% 
  sample_n(10)
```

```{r}
# Display structure of dataset
str(gdp)

# Convert Year to numeric 
# gdp$Year <- as.numeric(gdp$Year)

# Show the first few rows 
head(gdp)

# Summary statistics of numerical attributes
summary(gdp)

# Summary statistics of categorical attribute
table(gdp$`Indicator Code`)

# Check for missing values
sum(is.na(gdp))

# Check the dimension of data
dim(gdp)
```

```{r}
# Group the filtered data by Year and calculate the mean GDP for each year
mean_gdp_by_year <- gdp %>%
  group_by(Year) %>%
  summarize(mean_GDP = mean(GDP, na.rm = TRUE))

# Show the mean GDP for each year
print(mean_gdp_by_year)
```
Visualization for top 10 highest GDP by Counries Over Time.

```{r}
# Filter to include only the top 10 countries with the highest GDP values
top_10_gdp <- gdp %>%
  group_by(`Country Name`) %>%
  summarize(total_gdp = sum(GDP, na.rm = TRUE)) %>%
  top_n(10, total_gdp) %>%
  left_join(gdp, by = "Country Name")

# Convert GDP values to millions or billions
top_10_gdp <- top_10_gdp %>%
  mutate(GDP_formatted = case_when(
    GDP >= 1e12 ~ paste0(round(GDP / 1e12, 1), "T"),  # Convert to trillions
    GDP >= 1e9 ~ paste0(round(GDP / 1e9, 1), "B"),  # Convert to billions
    GDP >= 1e6 ~ paste0(round(GDP / 1e6, 1), "M"),   # Convert to millions
    TRUE ~ as.character(GDP)                        # Keep unchanged if less than 1 million
  ))

# Create a line plot of GDP over time with formatted values for the top 10 countries
ggplot(top_10_gdp, aes(x = Year, y = GDP, group = `Country Name`)) +
  geom_line(aes(color = `Country Name`)) +
  labs(title = "GDP Trends Over Time (Top 10 Countries)",
       x = "Year",
       y = "GDP",
       color = "Country") +
  scale_y_continuous(labels = function(x) paste0(x, "")) +  # Ensure y-axis labels are character type
  theme_minimal()
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


### Exploration of the Suicide Rate Dataset
```{r}
# Load data set from github
suicide_df = read_csv("https://raw.githubusercontent.com/Alexburk93/Data_Wrangling_EDA/main/data/raw_data/death-rate-from-suicides-gho%20new.csv")
```
```{r}
suicide_df$`Age-standardized suicide rate - Sex: both sexes` = as.double(suicide_df$`Age-standardized suicide rate - Sex: both sexes`)
summary(suicide_df)

# Consider removing Lesotho from the analysis 
suicide_df %>% 
  filter(is.na(`Age-standardized suicide rate - Sex: both sexes`))
```

```{r}
# Remove the Region / Income classes and Lesotho
suicide_df_countries = suicide_df %>% 
  filter(Code != "0", Code != "LSO")
```

```{r}
avg_country = suicide_df_countries %>% 
  group_by(Entity) %>% 
  summarise(avg_per_country = mean(`Age-standardized suicide rate - Sex: both sexes`)) %>% 
  arrange(desc(avg_per_country))
``` 

```{r}
ggplot(avg_country, aes(x = avg_per_country)) +
  geom_histogram() +
  labs(x = "Mean Suicide rate", y = "Frequency", title = "Distribution of mean of suicide rate") +
  theme_minimal()
```

### Exploration of the Happiness Dataset

```{r}
# Use the raw file URL
url <- "https://github.com/Alexburk93/Data_Wrangling_EDA/raw/main/data/raw_data/WHR20_DataForTable2.1.xls"
response <- GET(url)
content <- content(response, "raw")
temp <- tempfile(fileext = ".xls")
writeBin(content, temp)
happiness_df <- read_excel(temp)

# View the data
head(happiness_df)

```
```{r}
# create data frame with selected columns. Based on the description of the different variables.
happiness_df_filtered = happiness_df %>% 
  select(`Country name`, `year` , `Life Ladder`, `Social support`, `Healthy life expectancy at birth`, `Freedom to make life choices`, `Perceptions of corruption`)

summary(happiness_df_filtered)
```

```{r}
avg_happiness_per_country = happiness_df_filtered %>% 
  group_by(`Country name`) %>% 
  summarise(avg_happiness = mean(`Life Ladder`)) %>% 
  arrange(desc(avg_happiness))
```

```{r}
ggplot(avg_happiness_per_country, aes(x = avg_happiness)) +
  geom_histogram() +
  labs(x = "Mean Happiness Level", y = "Frequency", title = "Distribution of mean of happiness rate") +
  theme_minimal()
```

### Data for Final Presentation


Happiness Data for the 19 countries of interest from 2011 - 2019
```{r}
# Define the list of 19 countries
countries_of_interest <- c("Australia", "Belgium", "Brazil", "Canada", "Denmark",
                           "Finland", "France", "Germany", "Iceland", "Italy",
                           "Japan", "Netherlands", "New Zealand", "Norway", 
                           "South Africa", "Spain", "Sweden", "United Kingdom", 
                           "United States")

# Filter the data for the 19 countries
happiness_new<- happiness_df_filtered %>%
  filter(`Country name` %in% countries_of_interest)

# Define the date range of focus
start_year <- 2011
end_year <- 2020

# Filter the data for the date range
happiness_data <- happiness_new %>%
  filter(as.numeric(year) >= start_year & as.numeric(year) <= end_year)

# Show the filtered data
happiness_data %>% 
  sample_n(10)
```

```{r}
# Check if all countries of interest are present in the filtered data
missing_countries <- setdiff(countries_of_interest, unique(happiness_data$`Country name`))

# Print the missing countries, if any
if(length(missing_countries) > 0) {
  print("The following countries are missing from the filtered data:")
  print(missing_countries)
} else {
  print("All countries of interest are selected from the filtered data.")
}

```

```{r}
# Count the occurrences of each country in the filtered data
country_counts <- table(happiness_data$`Country name`)

# Print the country names and their counts
print("Country Name\t\tCount")
for (country in names(country_counts)) {
  cat(country, "\t\t\t", country_counts[country], "\n")
}

```

Suicide Data for the 19 countries of interest from 2011 - 2019
```{r}
# Change column name from `Entity` to `country name`
suicide_df <- rename(suicide_df, `Country name` = Entity)

# Define the list of 19 countries
countries_of_interest <- c("Australia", "Belgium", "Brazil", "Canada", "Denmark",
                           "Finland", "France", "Germany", "Iceland", "Italy",
                           "Japan", "Netherlands", "New Zealand", "Norway", 
                           "South Africa", "Spain", "Sweden", "United Kingdom", 
                           "United States")

# Filter the data for the 19 countries
new_suicide<- suicide_df %>%
  filter(`Country name` %in% countries_of_interest)

# Define the date range of focus
start_year <- 2011
end_year <- 2020

# Filter the data for the date range
suicide_data <- new_suicide %>%
  filter(as.numeric(Year) >= start_year & as.numeric(Year) <= end_year)

# Show the filtered data
suicide_data %>% 
  sample_n(10)
```

### Print all the data sets

```{r}
suicide_data %>% 
  sample_n(10)

happiness_data %>% 
  sample_n(10)

gdp %>% 
  sample_n(10)

bankruptcy_data %>% 
  sample_n(10)
```


### Label headers to align all the data set & view sample data
```{r}
gdp
bankruptcy_data
```


```{r}
# Change column name from `Entity` to `country name`
gdp_data <- rename(gdp, `Country name` = `Country Name`)
bankruptcy_data<- rename(bankruptcy_data, `Country name` = `Country`)
happiness_data = rename(happiness_data, `Year` = `year`)
suicide_data %>% 
  sample_n(10)

happiness_data %>% 
  sample_n(10)

gdp_data %>% 
  sample_n(10)

bankruptcy_data %>% 
  sample_n(10)
```
### Merging datasets

```{r}
gdp_data
happiness_data
bankruptcy_data
suicide_data
```

```{r}
# Convert "Year" column to character type in all datasets
suicide_data <- mutate(suicide_data, Year = as.character(Year))
happiness_data <- mutate(happiness_data, Year = as.character(Year))
gdp_data <- mutate(gdp_data, Year = as.character(Year))
bankruptcy_data <- mutate(bankruptcy_data, Year = as.character(Year))

# Merge the datasets
suicide_analysis <- suicide_data %>%
  left_join(happiness_data, by = c("Year", "Country name")) %>%
  left_join(gdp_data, by = c("Year", "Country name")) %>%
  left_join(bankruptcy_data, by = c("Year", "Country name"))

```
```{r}
# View sample data from the merged dataset
suicide_analysis 
```

Save the dataset in excel csv format to run analysis for the final project
```{r}

# Drop specified columns
suicide_analysis <- suicide_analysis %>%
  select(-c("Code", "COU", "PowerCode", "PowerCode Code"))

# Save the merged dataset as a CSV file
# Define the path to the folder on your desktop
desktop_path <- "/home/alex/Uni/Master_US/2_Semester/Class_Data_Wrangeling_EDA/Data_Wrangling_EDA/data/"

# Create the folder if it doesn't exist
dir.create(desktop_path, showWarnings = FALSE)

# Save the merged dataset as a CSV file in the specified folder
write.csv(suicide_analysis, file.path(desktop_path, "suicide_analysis_2.csv"), row.names = FALSE)
```


### References

**Title**: A Study for Effects of Economic Growth Rate and Unemployment Rate to Suicide Rate in Korea<br>
**Author**: Jong Soon Park, June Young Lee, Soon Duck Kim

**Objectives**:
We investigated the effects of the economic growth and unemployment rates on the suicide rate in Korea, between 1983 and 2000, using a time-series regression model. The purpose of this study was to model and test the magnitude of the rate of suicide, with the Korean unemployment rate and GDP. [2]

**Conclusion**: It was found that the suicide rate was closely related to the National's economic status of Korea, which is similar to the results found in studies in other countries. We expected, therefore, that this study could be used as the basis for further suicide-related studies. [2]

**Title**: Ten-year evolution of suicide rates and economic indicators in large Brazilian urban centers<br>
**Author**: Asevedo Elson, Ziebold Carolina, Diniz Elton, Gadelha Ary, Mari Jair

**Purpose of review**:
This was a retrospective ecological study to examine the relationship between suicide rates and economic indicators in large Brazilian urban centers. Data on macroeconomic indicators (GDP and unemployment rates) and suicide rates of the largest Brazilian cities were collected from January 2006 to December 2015. [3]

**Summary** 
The effect of economic indicators was heterogeneous among the centers, but, overall, the variation in suicide rates was inversely related to unemployment and did not show a significant relationship with GDP. These findings indicate a more complex link between economics and suicide whenever looking at local regional indicators. Further research should focus on possible intervening factors, what may inform better preventive interventions. [3]

**Title**: The association between nation-level social and economic indices and suicide rates: A pilot study <br>
**Author**: Ravi Philip Rajkumar

**Introduction**
Suicide is one of the leading causes of premature mortality worldwide. An analysis of global suicide data for the period 1990–2016, covering 195 countries, found that suicide was among the ten leading causes of years of life lost in Europe, the Americas and the Asia-Pacific region (Naghavi, 2019). Because of this, suicide prevention has been accorded one of the highest priorities in national and international public health plans, such as the World Health Organization's Comprehensive Mental Health Action Plan (World Health Organization, 2021). The act of suicide is a complex behavior which results from the interaction between an innate diathesis and external stressors (van Heeringen, 2012). [4]

**Conclusions**
Social factors are important determinants of suicide. Despite certain limitations, the results of the current study suggest that sustainable development may play a role in the mitigation of suicide risk, Economic inequality may contribute to variations in suicide risk, particularly in men. Moreover, the strength and direction of the associations between socioeconomic factors and suicide varies across income groups, highlighting the need for an in-depth understanding of each country's social, cultural and economic profile when planning large-scale social interventions aimed at suicide prevention. [4]

**Title**:Is Happiness the main cause for the rising suicide rate in the world? <br>
**Author**: Bong Jin Haw, Chong Yock Loon, Tai Hong Ting, Teoh Kah Chun, Wong Chin Hing

**Introduction**
This study investigates the impact of happiness on suicide rate. Beside theimpact of happiness, employment, gross domestic product, education, health expenditure and population on the suicide rate is also studied. The study divides the data into three sets including 31 developed and developing countries, 26 developed countries, and 5 developing countries. A significant relationship between happiness and the suicide rate is found in developing countries. [5]

**Conclusions**
Surprisingly, based on the result it can be concluded that the happiness in developing countries will significantly impact and influence the suicide rate indeed. It is important to realize that the happiness is negatively affect the suicide rate in developing countries which means that increase in the happiness will subsequently lower down the suicide rate. Moreover, other economic factors such as employment, gross domestic product, government health expenditure, education and population are also included in this study to carry out their impact towards the suicide rate. [5]

### Work Cited Source 
[1] https://www.iasp.info/wspd/references/<br>
[2] https://www.jpmph.org/journal/view.php?number=50 (visited 03/24/2024)<br>
[3] https://journals.lww.com/co-psychiatry/abstract/2018/05000/ten_year_evolution_of_suicide_rates_and_economic.15.aspx (visited 03/22/2024)<br>
[4] https://www.frontiersin.org/articles/10.3389/fsoc.2023.1123284/full<br>
[5] http://eprints.utar.edu.my/2869/1/fyp_BF_2018_BJH_-_1403312.pdf

